Abstract

### 2025-02-14 output/Main_text.Rmd
Version last response to reviewer 2
Version responses to reviewer 3 (WiP)

The representation of a latent space where the traits are correlated is more
accurate on a geometrical coordinate system with non-orthogonal axes
[@ackerman_multidimensional_2005-1] and (possibly) non-standard units,
often referred to as *general Cartesian coordinates* [@harman_modern_1970].
Such a space has the advantage that the basis vectors and the axes they yield
have a meaningful interpretation (albeit counterintuitive, as we will see)
in a multivariate statistical sense.
The general case of MIRT model estimation implies estimating possibly
non-standard latent dimensions and non-null covariances among them
(a more realistic case than an identity matrix).
Assuming orthonormality among traits is usually unrealistic, be it in the
cognitive [e.g., @reckase_multidimensional_2009] or the non-cognitive domain
[e.g., @thielmann_comparability_2022].
Moreover, certain applications impose the estimation of
non-identity latent covariance matrices (e.g., multigroup equating) even when
interpreting the latent dimensions as substantive constructs is not a goal
[for multidimensional linking and equating procedures,
see Chapter 9 in @reckase_multidimensional_2009].
As we will argue, the usual definitions of the multidimensional difficulty and
discrimination parameters do not generalize to the non-orthonormal geometry
necessary to accurately represent correlated latent spaces.

@@

Let's consider an orthonormal basis $`r LS_ORTH_BASIS_EQ`$ in $`r LATENT_SPACE`$
with the dot product as inner product.
Let $`r TRAIT_COORDS_EQ`$ and $`r TRAIT_COORDS_ORTH_EQ`$ be the coordinates of
any $`r TRAIT_VEC_IN_LS`$ in $`r LS_BASIS`$ and $`r LS_ORTH_BASIS`$,
respectively, such that $`r TRAIT_VECTOR_DEF_EQ`$.

@@

The (signed) distance to the origin,
given by the first component of $`r MIL_PARAM`$, has an interesting property:
By the definition of the inner product,
$`r MDISC_SYM`$ is strictly positive for non-null vectors.
Therefore, the sign of $`r DISTANCE_PARAM`$ is equal to
the sign of $`r INTERCEPT_PARAM`$.
The implication is that the item is displaced along
its measurement direction forwards or backwards with respect to the origin.
As the $`r MDISC_SYM`$ appears in the denominator of $`r DISTANCE_PARAM`$,
that displacement will be inversely proportional
to the discrimination of the item.
That is, the less discriminating the item is,
the further away it will be from the origin.

>>>>

The representation of a latent space where the traits are correlated is more
accurate on a geometrical coordinate system with non-orthogonal axes
[@ackerman_multidimensional_2005-1, p. 16, see footnote 1]
and (possibly) non-standard units,
often referred to as *general Cartesian coordinates* [@harman_modern_1970].
Such a space has the advantage that the basis vectors and the axes they yield
have a meaningful interpretation (albeit counterintuitive, as we will see)
in a multivariate statistical sense.
The general case of MIRT model estimation implies estimating possibly
non-standard latent dimensions and non-null covariances among them
(a more realistic case than an identity matrix).
Assuming orthonormality among traits is usually unrealistic, be it in the
cognitive [e.g., @reckase_multidimensional_2009] or the non-cognitive domain
[e.g., @thielmann_comparability_2022].
Moreover, certain applications impose the estimation of
non-identity latent covariance matrices (e.g., multigroup equating) even when
interpreting the latent dimensions as substantive constructs is not a goal
[for multidimensional linking and equating procedures,
see Chapter 9 in @reckase_multidimensional_2009].
As we will argue, the usual definitions of the multidimensional difficulty and
discrimination parameters do not generalize to the non-orthonormal geometry
necessary to accurately represent correlated latent spaces.

@@

Let's consider an orthonormal basis $`r LS_ORTH_BASIS_EQ`$ in $`r LATENT_SPACE`$
with the dot product as inner product.
Let $`r TRAIT_COORDS_EQ`$ and $`r TRAIT_COORDS_ORTH_EQ`$ be the coordinates of
any $`r TRAIT_VEC_IN_LS`$ in $`r LS_BASIS`$ and $`r LS_ORTH_BASIS`$,
respectively, such that $`r TRAIT_VECTOR_DEF_EQ`$.
That is, vector $`r TRAIT_VECTOR`$ is represented in each of the two bases by
the linear combination of its element vectors
scaled by the coordinates of $`r TRAIT_VECTOR`$ in that basis.

@@

The (signed) distance to the origin,
given by the first component of $`r MIL_PARAM`$, has an interesting property:
By the definition of the inner product,
$`r MDISC_SYM`$ is strictly positive for non-null vectors.
Therefore, the sign of $`r DISTANCE_PARAM`$ is equal to
the sign of $`r INTERCEPT_PARAM`$.
The implication is that the item is displaced along
its measurement direction forwards or backwards with respect to the origin.
As the $`r MDISC_SYM`$ appears in the denominator of $`r DISTANCE_PARAM`$,
that displacement will be inversely proportional
to the discrimination of the item.
That is, the less discriminating the item is,
the further away it will be from the origin.
Of course, it is also directly proportional to
the intercept parameter $`r INTERCEPT_PARAM`$.
This implies that, if $`r INTERCEPT_NULL_EQ`$, $`r DISTANCE_NULL_EQ`$ as well,
regardless of the discrimination.
